Testing the motor and cognitive foundations of Paleolithic social transmission \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2021-07-31 manuscript/Testing-the-cognitive-foundations-of-Paleolithic-social-transmission.tex
update the text to the latest google doc version

Stone tools have long been seen as a key source of evidence for
understanding human behavioral and cognitive evolution (Darwin 1871;
Oakley 1949; Washburn 1960). Pathbreaking attempts to infer specific
cognitive capacities from this evidence largely focused on the basic
requirements of tool production (Isaac 1976; Wynn 1979; Gowlett 1984;
Wynn and Coolidge 2004). More recently, increasing attention has been
directed to the processes and demands of stone tool making skill
acquisition (Roux, Bril, and Dietrich 1995; Stout 2002; Stout et al.
2005; Geribàs, Mosquera, and Vergès 2010; Nonaka, Bril, and Rein 2010;
Stout et al. 2011; Putt, Woods, and Franciscus 2014; Hecht, Gutman,
Khreisheh, et al. 2015; Duke and Pargeter 2015; Morgan et al. 2015;
Stout and Khreisheh 2015; Lombao, Guardiola, and Mosquera 2017; Putt et
al. 2017; Cataldo, Migliano, and Vinicius 2018; Putt, Wijeakumar, and
Spencer 2019; Pargeter and Shea 2019; Pargeter et al. 2020). This is
motivated by the expectation that the most salient cognitive demands of
tool making should occur during learning rather than routine expert
performance (Stout and Khreisheh 2015) and by interest in the relevance
of different social learning mechanisms such as imitation (Rein, Nonaka,
and Bril 2014; Stout et al. 2019), emulation (Tehrani and Riede 2008;
Wilkins 2018), and language (Ohnuma, Aoki, and Akazawa 1997; Putt,
Woods, and Franciscus 2014; Morgan et al. 2015; Lombao, Guardiola, and
Mosquera 2017; Putt et al. 2017; Cataldo, Migliano, and Vinicius 2018)
to the reproduction of Paleolithic technologies.

Studies investigating these questions have used a range of different
experimental designs (e.g., varying technological goals/instructions,
training times, raw materials, live vs.~recorded instruction,
lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached
disparate conclusions regarding the neurocognitive and social
foundations of skill acquisition. It is plausible that these discordant
results reflect actual diversity in how humans acquire and master stone
tool making skills. However, this failure of results to generalize
across artificial experimental manipulations
(\href{cf.\%20Yarkoni\%202020}{cf}. Yarkoni 2020) also raises doubts
regarding the external validity (Eren et al. 2016) of conclusions with
respect to real-world Paleolithic learning contexts. To address this, we
conducted an exploratory study that draws on lessons from previous
research in an attempt to balance the pragmatic and theoretical
tradeoffs inherent in experimental studies of stone knapping skill
acquisition (Pargeter and Shea 2019; Stout and Khreisheh 2015).

@@

We focus on early stage learning because it has been found to be
relatively rapid, variable across individuals, and predictive of later
outcomes (Pargeter and Shea 2019; Stout and Khreisheh 2015; Putt,
Wijeakumar, and Spencer 2019), and thus provides a reasonable
expectation of generating meaningful data on skill and learning
variation while minimizing training costs. Moreover, understanding the
minimum training times necessary to detect changes in tool making skill
will help archaeologists design more realistic and cost-effective
experiments. To further manage costs, we limited our study to only two
learning conditions (observation only vs.~active teaching). This targets
a key controversy in human evolution, namely the origins of teaching and
language (Gärdenfors and Högberg 2017; Morgan et al. 2015), while
avoiding highly artificial manipulations of dubious relevance to
real-world Paleolithic learning. These choices allowed us to invest more
in other aspects of research design that we identified as theoretically
important, including measurement of individual differences in cognition
and behavior, inclusion of an in-person, fully interactive teaching
condition, and use of naturalistic raw materials. Sample size remained
small in this internally funded exploratory study but could easily be
scaled up at funding levels typical of pre- and post-doctoral research
grants in archaeology.

@@

To date, a small number of ``neuroarchaeological'' studies have reported
associations between individual knapping performance and brain structure
or physiological responses. Hecht et al. (2015) reported
training-related changes in white matter integrity (fractional
anisotropy {[}FA{]}) that correlated with individual differences in
practice time and striking accuracy change. The regional patterning of
FA changes also varied across individuals, with only those individuals
who displayed early increases in FA under the right ventral precentral
gyrus (premotor cortex involved in movement planning and guidance)
showing striking accuracy improvement over the training period. Putt et
al. (2019) similarly found that the proportion of flakes to shatter
produced by individuals during handaxe making correlated with dorsal
precentral gyrus (motor cortex) activation. Pargeter et al. (2020) used
a flake prediction paradigm (modeled after Nonaka, Bril, and Rein 2010)
to confirm that striking force and accuracy are important determinants
of handaxe-making success. These findings all point to the central role
of perceptual-motor systems (Stout and Chaminade 2007) and coordination
(Roux, Bril, and Dietrich 1995) in knapping skill acquisition. In
addition, Putt et al. (2019) also found successful flake production to
be associated with prefrontal (working memory/cognitive control)
activation and Stout et al.~(2015) found that prefrontal activation
correlated with success at a strategic judgement (platform selection)
task which in turn was predictive of success at out-of-scanner handaxe
production. Such investigations are thus starting to chart out the more
specific contributions of different neural systems to particular aspects
of knapping skill acquisition. To date, however, the
cognitive/functional interpretation of systems identified in this manner
has largely relied on informal reverse inference (reasoning backward
from observed activations to inferred mental processes) from published
studies of other tasks that activated the same regions, an approach
which is widely regarded as problematic (Poldrack 2011).

Here we take a more direct, psychometric approach to measuring
individual differences in perceptual-motor coordination and cognition.
Psychometric instruments (e.g., tasks, questionnaires) are designed to
assess variation in cognitive traits and states, such as fluid
intelligence, working memory, attention, motivation, and personality,
that have been of theoretical interest to cognitive archaeologists
(e.g., Wynn and Coolidge 2016). It is thus surprising that they have
been almost entirely neglected in experimental studies of knapping
skill. In the only published example we are aware of, Pargeter et al.
(2019) reported significant effects of variation in planning and problem
solving (Tower of London test (Shallice, Broadbent, and Weiskrantz
1982)) and cognitive set shifting (Wisconsin Card Sort test (Grant and
Berg 1948)) on early stage handaxe learning. Of course, cognition is not
the only thing that can affect knapping performance. Flake prediction
experiments highlight the importance of regulating movement
speed/accuracy trade-offs (Nonaka, Bril, and Rein 2010; Pargeter et al.
2020) and studies of muscle recruitment (Marzke et al. 1998) and manual
pressure (Williams-Hatala et al. 2018; Alastair J. M. Key and Dunmore
2018) during knapping highlight basic strength requirements. Along these
lines, Key and Lycett (2019) found that individual differences in hand
size, shape, and especially grip strength were better predictors of
force loading during stone tool use than were attributes of the tools
themselves. However, we are unaware of any such studies of biometric
influences on variation in knapping success. Finally, the time and
effort demands of knapping skill acquisition suggest that differences in
personality (e.g., self-control and ``grit'' (Pargeter and Shea 2019),
motivation (Stout 2002), and social vs.~individual learning strategies
(Miu et al. 2020) might also affect learning outcomes. We are again
unaware of any previous studies that have assessed such effects. In this
study, we assessed all participants with a battery of tests including
grip strength, movement speed/accuracy, spatial working memory, fluid
intelligence, self-control, tendency to use social information, and
motivation/engagement with the tool making task. We were particularly
interested in the possibility that these variables might not only impact
learning generally, but might also have different effects under
different learning conditions.~

@@

@@

@@

@@

@@

@@

@@

\hypertarget{lithic-analysis}{%
\subsection{\texorpdfstring{\textbf{Lithic
Analysis}}{Lithic Analysis}}\label{lithic-analysis}}

@@

>>>>

Stone tools have long been seen as a key source of evidence for
understanding human behavioral and cognitive evolution (Darwin 1871;
Oakley 1949; Washburn 1960). Pathbreaking attempts to infer specific
cognitive capacities from this evidence largely focused on the basic
requirements of tool production (Isaac 1976; Wynn 1979; Gowlett 1984;
Wynn and Coolidge 2004). More recently, increasing attention has been
directed to the processes and demands of stone tool making skill
acquisition (Roux, Bril, and Dietrich 1995; Stout 2002; Stout et al.
2005; Geribàs, Mosquera, and Vergès 2010; Nonaka, Bril, and Rein 2010;
Stout et al. 2011; Putt, Woods, and Franciscus 2014; Hecht, Gutman,
Khreisheh, et al. 2015; Duke and Pargeter 2015; Morgan et al. 2015;
Stout and Khreisheh 2015; Lombao, Guardiola, and Mosquera 2017; Putt et
al. 2017; Cataldo, Migliano, and Vinicius 2018; Putt, Wijeakumar, and
Spencer 2019; Pargeter, Khreisheh, and Stout 2019; Pargeter et al.
2020). This is motivated by the expectation that the most salient
cognitive demands of tool making should occur during learning rather
than routine expert performance (Stout and Khreisheh 2015) and by
interest in the relevance of different social learning mechanisms such
as imitation (Rein, Nonaka, and Bril 2014; Stout et al. 2019), emulation
(Tehrani and Riede 2008; Wilkins 2018), and language (Ohnuma, Aoki, and
Akazawa 1997; Putt, Woods, and Franciscus 2014; Morgan et al. 2015;
Lombao, Guardiola, and Mosquera 2017; Putt et al. 2017; Cataldo,
Migliano, and Vinicius 2018) to the reproduction of Paleolithic
technologies.

Studies investigating these questions have used a range of different
experimental designs (e.g., varying technological goals/instructions,
training times, raw materials, live vs.~recorded instruction,
lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached
disparate conclusions regarding the neurocognitive and social
foundations of skill acquisition. It is plausible that these discordant
results reflect actual diversity in how humans acquire and master stone
tool making skills. However, this failure of results to generalize
across artificial experimental manipulations
(\href{cf.\%20Yarkoni\%202020}{cf}. Yarkoni 2020) also raises doubts
regarding the external validity (Eren et al. 2016) of conclusions with
respect to real-world Paleolithic learning contexts. To address this, we
conducted an exploratory study that draws on lessons from previous
research in an attempt to balance the pragmatic and theoretical
tradeoffs inherent in experimental studies of stone knapping skill
acquisition (Pargeter, Khreisheh, and Stout 2019; Stout and Khreisheh
2015).

@@

We focus on early stage learning because it has been found to be
relatively rapid, variable across individuals, and predictive of later
outcomes (Pargeter, Khreisheh, and Stout 2019; Stout and Khreisheh 2015;
Putt, Wijeakumar, and Spencer 2019), and thus provides a reasonable
expectation of generating meaningful data on skill and learning
variation while minimizing training costs. Moreover, understanding the
minimum training times necessary to detect changes in tool making skill
will help archaeologists design more realistic and cost-effective
experiments. To further manage costs, we limited our study to only two
learning conditions (observation only vs.~active teaching). This targets
a key controversy in human evolution, namely the origins of teaching and
language (Gärdenfors and Högberg 2017; Morgan et al. 2015), while
avoiding highly artificial manipulations of dubious relevance to
real-world Paleolithic learning. These choices allowed us to invest more
in other aspects of research design that we identified as theoretically
important, including measurement of individual differences in cognition
and behavior, inclusion of an in-person, fully interactive teaching
condition, and use of naturalistic raw materials. Sample size remained
small in this internally funded exploratory study but could easily be
scaled up at funding levels typical of pre- and post-doctoral research
grants in archaeology.

@@

To date, a small number of ``neuroarchaeological'' studies have reported
associations between individual knapping performance and brain structure
or physiological responses. Hecht et al. (2015) reported
training-related changes in white matter integrity (fractional
anisotropy {[}FA{]}) that correlated with individual differences in
practice time and striking accuracy change. The regional patterning of
FA changes also varied across individuals, with only those individuals
who displayed early increases in FA under the right ventral precentral
gyrus (premotor cortex involved in movement planning and guidance)
showing striking accuracy improvement over the training period. Putt et
al. (2019) similarly found that the proportion of flakes to shatter
produced by individuals during handaxe making correlated with dorsal
precentral gyrus (motor cortex) activation. Pargeter et al. (2020) used
a flake prediction paradigm (modeled after Nonaka, Bril, and Rein 2010)
to confirm that striking force and accuracy are important determinants
of handaxe-making success. These findings all point to the central role
of perceptual-motor systems (Stout and Chaminade 2007) and coordination
(Roux, Bril, and Dietrich 1995) in knapping skill acquisition. In
addition, Putt et al. (2019) also found successful flake production to
be associated with prefrontal (working memory/cognitive control)
activation and Stout et al. (2015) found that prefrontal activation
correlated with success at a strategic judgement (platform selection)
task which in turn was predictive of success at out-of-scanner handaxe
production. Such investigations are thus starting to chart out the more
specific contributions of different neural systems to particular aspects
of knapping skill acquisition. To date, however, the
cognitive/functional interpretation of systems identified in this manner
has largely relied on informal reverse inference (reasoning backward
from observed activations to inferred mental processes) from published
studies of other tasks that activated the same regions, an approach
which is widely regarded as problematic (Poldrack 2011).

Here we take a more direct, psychometric approach to measuring
individual differences in perceptual-motor coordination and cognition.
Psychometric instruments (e.g., tasks, questionnaires) are designed to
assess variation in cognitive traits and states, such as fluid
intelligence, working memory, attention, motivation, and personality,
that have been of theoretical interest to cognitive archaeologists
(e.g., Wynn and Coolidge 2016). It is thus surprising that they have
been almost entirely neglected in experimental studies of knapping
skill. In the only published example we are aware of, Pargeter et al.
(2019) reported significant effects of variation in planning and problem
solving (Tower of London test (Shallice, Broadbent, and Weiskrantz
1982)) and cognitive set shifting (Wisconsin Card Sort test (Grant and
Berg 1948)) on early stage handaxe learning. Of course, cognition is not
the only thing that can affect knapping performance. Flake prediction
experiments highlight the importance of regulating movement
speed/accuracy trade-offs (Nonaka, Bril, and Rein 2010; Pargeter et al.
2020) and studies of muscle recruitment (Marzke et al. 1998) and manual
pressure (Williams-Hatala et al. 2018; Alastair J. M. Key and Dunmore
2018) during knapping highlight basic strength requirements. Along these
lines, Key and Lycett (2019) found that individual differences in hand
size, shape, and especially grip strength were better predictors of
force loading during stone tool use than were attributes of the tools
themselves. However, we are unaware of any such studies of biometric
influences on variation in knapping success. Finally, the time and
effort demands of knapping skill acquisition suggest that differences in
personality (e.g., self-control and ``grit'' (Pargeter, Khreisheh, and
Stout 2019), motivation (Stout 2002), and social vs.~individual learning
strategies (Miu et al. 2020) might also affect learning outcomes. We are
again unaware of any previous studies that have assessed such effects.
In this study, we assessed all participants with a battery of tests
including grip strength, movement speed/accuracy, spatial working
memory, fluid intelligence, self-control, tendency to use social
information, and motivation/engagement with the tool making task. We
were particularly interested in the possibility that these variables
might not only impact learning generally, but might also have different
effects under different learning conditions.~

@@

Toth and Schick (1993) were, however, careful to point out that extinct
hominid learning strategies and capacities might differ from modern
experimental participants. Even leaving aside potential species
differences in social learning (cf. Morgan et al. 2015; Stout et al.
2019), reliance on explicit verbal instruction varies widely across
modern human societies (e.g., Boyette and Hewlett 2017). The WEIRD
(Western, educated, industrialized, rich, democratic (Henrich, Heine,
and Norenzayan 2010)) teachers and learners typical of knapping
experiments arguably represent an extreme bias toward such instruction.
Simply instructing such participants not to speak during an experiment
(or to demonstrate but not gesture, etc. (Morgan et al. 2015)) is likely
to underestimate the efficacy of non-verbal teaching and learning in
cultural contexts where it is more common, let alone in a hypothetical
pre-linguistic hominid species.

Such concerns are exacerbated in experiments using pre-recorded
instructional videos or extremely short training periods. Video does not
allow the interactive teaching that is favored even in formal academic
knapping classes (e.g., Shea 2015) and is almost certainly typical of
traditional learning contexts (e.g., Stout 2002). It is not known how
video presentation affects the efficacy of teaching generally, or the
relative effectiveness of different forms of instruction. Going further,
some experiments have manipulated the presence/absence of verbal
instruction by presenting the same video with and without sound (Putt et
al. 2017) or the sound track without the video (Cataldo, Migliano, and
Vinicius 2018). While this provides experimental control, it does not
allow the instructor to adjust their multi-modal (Levinson and Holler
2014) communication strategies as they would naturally do, for example
through pointing and pantomime. To simply remove a communication channel
without allowing any such adaptation is highly artificial and risks
generating results that cannot be generalized beyond the specific
context of the experiment (Yarkoni 2020). Similarly, unnaturally short
training periods (e.g., 5-15 minutes (Morgan et al. 2015; Lombao,
Guardiola, and Mosquera 2017)) might misrepresent the relative efficacy
of different teaching strategies under more realistic conditions (Whiten
2015; Stout and Khreisheh 2015). Even the longest training times to date
(Stout and Khreisheh 2015; Pargeter, Khreisheh, and Stout 2019) have not
produced knapping skills comparable to relevant archaeological examples,
and were achieved by limiting sample size and using only one teaching
condition.

For these reasons, we sought to explore a middle path between
experimental expedience and realism by limiting our experiment to two
relatively naturalistic learning conditions and a moderate learning
period of two hours. As in previous experiments (Stout et al. 2011;
Hecht, Gutman, Khreisheh, et al. 2015; Pargeter, Khreisheh, and Stout
2019) the first condition was unrestricted, interactive instruction in
small groups, essentially reproducing the ``natural'' teaching/learning
context familiar (cf. Shea 2015) to our WEIRD instructor and student
participants. The second condition allowed observation only, with the
experimenter visible making flakes but not interacting in any way with
learners. This absence of teaching is again a familiar social context
for our participants and did not require any novel behaviors from the
instructor. It matches the ``imitation/emulation'' condition of Morgan
et al. (2015) although we make no assumptions regarding learning
mechanisms. We did not include a ``reverse engineering'' or ``end-state
emulation'' condition in which only finished products were visible. This
has been advocated as an important baseline or control condition (Whiten
2015) to distinguish observational from individual learning, but is not
likely to model any typical Paleolithic learning context nor to stand as
an adequate proxy for the cognition of hominid species with different
social learning capacities. There is no reason to assume neurocognitive
and behavioral processes of reverse-engineering problem solving in
modern humans (e.g., Allen, Smith, and Tenenbaum 2020) approximate the
social learning processes of hominids with more ape-like action
observation/imitation capacities (Hecht, Gutman, et al. 2013; Hecht,
Murphy, et al. 2013; Stout et al. 2019).

We selected a two-hour learning period for both pragmatic and
theoretical reasons. Pargeter et al. (Pargeter, Khreisheh, and Stout
2019) found that even \textasciitilde90 hours of fully interactive
instruction and practice was insufficient to achieve handaxe-making
skills comparable to the later Acheulean site of Boxgrove
(García-Medrano et al. 2019; Stout et al. 2014), and estimated actual
time to mastery as ranging from 121 to 441 hours for different
participants. However, they observed the greatest, fastest, and most
individually variable skill increases during the first 20 hours of
practice. In addition, initial performance was moderately correlated
with later achievement. This suggests that studying early-stage learning
may be a pragmatic alternative, especially for research investigating
individual differences in aptitude. Studies of simple flake production
similarly document large initial variation (Stout and Khreisheh 2015)
and rapid early progress (Stout and Khreisheh 2015; Putt, Wijeakumar,
and Spencer 2019; Stout and Semaw 2006). We designed the current study
to test the utility of studying learning and variation during the first
two hours of simple flaking instruction/practice, in hopes of finding a
viable compromise between experimental realism and cost

Lithic raw materials vary in size, shape, and fracture mechanical
properties that affect the difficulty of achieving different knapping
goals (Eren et al. 2014). Unfortunately, it can be difficult and/or
expensive to procure authentic raw materials. Experimental studies of
knapping skill have often used proxy materials such as flint (Morgan et
al. 2015; Nonaka, Bril, and Rein 2010; Cataldo, Migliano, and Vinicius
2018), limestone (Stout and Semaw 2006), porcelain (Khreisheh, Davies,
and Bradley 2013), or heat-treated chert (Putt, Woods, and Franciscus
2014; Putt et al. 2017; Putt, Wijeakumar, and Spencer 2019)to model
Oldowan and early Acheulean technologies executed in other materials. As
well as being more readily available, these proxies are generally easier
to knap. This has the benefit of reducing required practice time, but it
is unclear how it might affect learning demands more generally or the
efficacy of different learning conditions/strategies specifically.

To address this, some studies have attempted to more closely match
experimental and archaeological raw material types (Stout et al. 2011;
Duke and Pargeter 2015; Pargeter, Khreisheh, and Stout 2019). However,
raw materials vary across individual clasts within as well as between
types. This has led to interest in standardizing experimental core
morphology (Nonaka, Bril, and Rein 2010) and composition, even if this
means using artificial materials such as porcelain (Khreisheh, Davies,
and Bradley 2013), brick (Geribàs, Mosquera, and Vergès 2010; Lombao,
Guardiola, and Mosquera 2017), or foam blocks (Schillinger, Mesoudi, and
Lycett 2014). Such manipulations enhance experimental control and
internal validity (Eren et al. 2016) at the expense of external
generalizability to actual archaeological conditions. Specifically, they
allow more robust results from smaller samples but eliminate a core
element of real-world knapping skill: the ability to produce consistent
results from variable materials (Pelegrin 1990; Stout 2013). For
example, Pargeter et al. (2020) found that predicting specific flaking
outcomes on actual handaxe preforms was both more difficult and less
technologically important than expected from previous work with
standardized, frustum-shaped cores (Nonaka, Bril, and Rein 2010). The
alternative to control is to incorporate raw material size, shape, and
composition as experimental variables (e.g., Stout et al. 2019). This
allows consideration of raw material selection and response to variation
as aspects of skill but correspondingly increases the sample sizes
required to identify patterning. In considering these issues, we again
chose to explore a middle path between pragmatism and realism by
employing commercially purchased basalt similar to that known from East
African Oldowan sites, allowing clast size and shape to vary within set
limits, and selecting the particular clasts provided to each participant
to approximate the same distribution.

@@

This research was approved by the Emory Institutional Review Board
(IRB00113024). All participants provided written informed consent and
completed a video release form
(\url{https://databrary.org/support/irb/release-template.html}).

@@

Participants were asked to visit the Paleolithic Technology Lab at Emory
University to complete one three-hour session. Participants were
scheduled to attend in six groups of four, however one of these groups
had only three participants due to a no-show on the day of the
experiment. Each visit began with the collection of individual
differences measures, which took approximately one hour. After that,
participants undertook 105 minutes (two hours minus a 15-minute break
after 1 hour) of stone tool making practice. This session was
video-recorded, and all lithic products were collected. After the tool
making task, participants completed an ``exit questionnaire'' comprising
the Intrinsic Motivation Inventory (see below).

Participants were compensated for their time with a \$30 gift card. They
also had the opportunity to earn a performance bonus of \$5, \$10, \$15
or \$20 on the gift card. They were told that this bonus would depend on
``how well they did'' on the last core of their practice session. The
actual performance measure was not specified, but in order to allow on
the spot payment a simple measure of the percentage of starting weight
removed from the final core was used such that: \textgreater{} 30\%
earned \$5, \textgreater{} 40\% earned \$10, \textgreater{} 50\% earned
\$15, \textgreater{} 75\% earned \$20.

@@

We used five individual difference measures for this study:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Grip strength was measured in kilograms using an electronic hand
  dynamometer (Camry EH101). Strength was measured twice and the higher
  value recorded. Grip strength is a simple measure that is well
  correlated with overall muscular strength (Wind et al. 2010) and a
  range of other health and fitness measures (Sasaki et al. 2007). It is
  hypothesized to be relevant to generating kinetic energy for fracture
  initiation (Nonaka, Bril, and Rein 2010) as well as control and
  support of the hammerstone (Williams-Hatala et al. 2018) and core
  (Faisal et al. 2010; Alastair J. M. Key and Dunmore 2015).
\item
  Motor accuracy was assessed using a ``Fitts Law'' reciprocal tapping
  task. Fitts Law describes the trade-off between speed and accuracy in
  human movement, classically measured by tapping back and forth between
  two targets of varying size and spacing (Fitts 1954). Archaeologists
  have proposed (Stout 2002; Pargeter et al. 2020) that management of
  this trade-off is critical to the accurate application of appropriate
  force seen in skilled knapping (Nonaka, Bril, and Rein 2010; Roux,
  Bril, and Dietrich 1995). We implemented this test on a Surface Pro
  tablet running free software (FittsStudy Version 4.2.8, default
  settings) developed by the Accessible Computing Experiences lab (Jacob
  O. Wobbrock, director) at the University of Washington
  (\url{depts.washington.edu/acelab/proj/fittsstudy/index.html}).
  Participants use a touchscreen pen to tap between ribbons on the
  screen, with average movement time as the performance metric.
\item
  Visuospatial working memory is the capacity to ``hold in mind,'' which
  researchers have hypothesized to be important in stone toolmaking
  performance (Coolidge and Wynn 2005). It also might support a learning
  process known as `chunking,' in which multiple items or operations are
  combined into summary chunks stored in long term memory, that is
  thought to be important in the acquisition of knapping and other
  skills (Pargeter, Khreisheh, and Stout 2019). We measured visuospatial
  working memory using a free n-back task
  (\href{https://wmp.education.uci.edu/software/}{wmp.education.uci.edu/software/})
  developed by the Working Memory and Plasticity Laboratory at the
  University of California, Irvine (Susanne Jaeggi, PI) and implemented
  in E-Prime software on a desktop computer. In this task, participants
  are asked to remember the position of blue squares presented
  sequentially on the screen and touch a key when the current position
  matches that 1, 2, 3\ldots n iterations back. Progression to blocks
  with increasing values of n is contingent on exceeding a threshold
  success rate. Performance was measured as the highest n achieved.
\item
  Fluid intelligence (Cattell 1963) refers to the capacity to engage in
  abstract reasoning and problem solving in a way that is minimally
  dependent on prior experience. It complements ``crystallized
  intelligence'' (the ability to apply learned procedures and knowledge)
  as one of the two factors (gf, gc) comprising so-called ``general
  intelligence'' (g). Fluid intelligence is closely related to the
  executive control of attention and manipulation of information held in
  working memory (Engle 2018)(Engle 2018). It is hypothesized to support
  technological innovation (Coolidge and Wynn 2005) and/or the
  intentional learning of new skills (Unsworth and Engle 2005; Stout and
  Khreisheh 2015). We measured fluid intelligence using the short
  version (Bilker et al. 2012) of the classic Raven Progressive Matrices
  task, which requires participants to complete increasingly difficult
  pattern matching questions.~~
\item
  The use of social information for learning and decision making varies
  across individuals and societies (Molleman, Kurvers, and van den Bos
  2019). Such variation is a key topic for understanding social learning
  and cultural evolutionary processes (Kendal et al. 2018; Heyes 2018;
  Miu et al. 2020) and represents a potential confound for assessing
  experimental effects of different social learning conditions. We
  measured participants' tendency to rely on social information
  vs.~their own insights using the Berlin Estimate AdjuStment Task
  (BEAST) developed by Molleman et al. (2019). In this task,
  participants are present with large arrays of items on a screen and
  asked to estimate the number present. They are then provided with
  another person's estimate and allowed to provide a second estimate.
  The participants' average adjustment between first and second
  estimates provides a measure of their propensity to rely on social
  information.
\end{enumerate}

@@

After individual difference testing, participants engaged in a 2-hour
stone tool making session, with a 15-minute break after 1 hour.
Participants were instructed not to seek out additional training or
information on stone tool making (i.e., via the internet) during these
breaks. Each group of participants was randomly assigned to one of two
experimental conditions: no teaching or teaching. In both conditions,
participants were first given an opportunity to inspect and handle
examples (\textbf{Figure}) of the kind of stone tools (flakes) they are
being asked to produce. They were told that their objective was to
produce as many flakes as possible from the materials provided. This
meant that even the untaught condition included some minimal instruction
(being told the objective) , however this was considered to be
unavoidable without creating a much more elaborate and naturalistic
context in which participants would develop their own technological
goals. Such a design would also be expected to increase behavioral
variability, demanding correspondingly larger samples of participants to
identify patterns and making direct comparisons with the taught
condition.

@@

Each participant was provided with 9 cores for use over the 2-hour
experiment. These cores were produced from larger chunks of a
fine-grained basalt purchased from neolithics.com by fracturing them
with a sledgehammer. This produced irregular, angular chunks for use in
the experiment, weighing between 459g - 1876g (mean = 975g). All cores
were weighed, measured (Length, Width, Thickness), and painted white so
that new fracture surfaces could be discriminated from those created
during production. Cores were sorted by shape and weight and then
distributed evenly to each participant. As a result, there were no
significant difference across participants in the mean weight (ANOVA, df
= 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04,
df1=22, df2 = 184, p = 0.4) or shape (Length x Width/Thickness: ANOVA,
df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p =
0.9) of cores provided. This was also true comparing the two
experimental conditions (Taught vs.~Untaught mean weight = 1001g
vs.~956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4;
mean shape = 221.43 vs.~221.45, t = -0.003, df = 205, p = 0.9, Levene's
Test F = 3.8, p = 0.05). Participants were, however, allowed to choose
which cores to work on so that differences in the weight and shape of
cores actually used across participants and conditions could still
emerge as a result of selection bias.

Sixty pounds of 3-to-5 inch basalt ``Mexican Beach Pebbles'' were
purchased from a landscaping supply company for use as hammerstones in
the experiment. Of these, 90 were selected as suitable for use. These
weighed between 213g-1360g (mean = 425) and varied in elongation (L/W =
1.01 to 2.65) and relative thickness (LxW/T = 90.48 to 283.67).
Forty-five stones were placed in the middle of the knapping area (Figure
2) for participants to freely choose from during the experiment. Broken
hammerstones were replaced from the reserve to maintain a consistent
number and range of choices. Each hammerstone was numbered and
participants' choices were recorded along with the number of the core(s)
being worked on with a particular hammerstone.\\

@@

In both conditions, three researchers were present to record activities
and collect materials. Participants were seated in a circle (Shea 2015)
and experiments were video recorded using two cameras (Figure 1).
Participants were free to select hammerstones from the common pile and
to work on any or all of their nine assigned cores in any order they
preferred. However, each core and all associated debitage were collected
before participants were allowed to start working on a new core, so it
was not possible to partially work and then return to a particular core
later. The order of cores used and associated hammerstones were recorded
for each participant during the experiment.

In the untaught condition, a researcher (DS) sat with the participants
and made stone tools but remained silent and made no effort to
facilitate learning (e.g., through gesture, modified performance, facial
expression, attention direction, or verbal instruction). Over the 2-hour
period, the researcher completely reduced four cores (one every
\textasciitilde30 minutes). Participants were not restricted from
talking to each other, as this would create an unnatural and potentially
stressful social context that might affect learning. Participants were
asked to avoid any form of communication about the tool making task
specifically, and they complied with this request. Participants in this
condition thus had the opportunity to observe tool making by an expert
and/or by other learners, should they choose to do so, but received no
intentional instruction.

In the Taught condition, there were no restrictions on participant
interaction and the researcher engaged in direct active teaching (Kline
2015) of tool-making techniques through verbal instruction,
demonstration, gesture, and shaping of behavior. The instructor has a
moderate level of experience teaching basic knapping skills to students
in undergraduate archaeology classes and to participants in previous
knapping research (e.g., Stout et al. 2011). The pedagogical strategy
employed was based on the instructor's own learning experiences and
theoretical interpretations (e.g., Pargeter et al. 2020), and focused on
coaching participants in effective body postures, movement patterns, and
grips as well as the assessment of viable core morphology.~ ~

\textbf{Lithic Analysis}

All finished cores were weighed and measured (L, W, T). Delta weight was
calculated as (Start weight-End weight)/Start weight. All detached
pieces (DPs) were collected and weighed. We did not sort DPs into types
(e.g., whole flakes, fragments) as this would have greatly increased
processing time and it is not clear that such distinctions add relevant
information regarding utility/desirability beyond that supplied by
metrics (Stout et al. 2019). All DPs larger than 40mm in maximum
dimension were photographed and measured. It is conventional in Early
Stone Age lithic analysis to employ a 20 mm cut-off. We selected a
higher threshold for both pragmatic (analysis time) and theoretical
reasons. Flake use experiments have shown that flakes weighing less than
5--10 g or with a surface area below 7--10 cm2 (Prasciunas 2007) or with
a maximum dimension \textless50-60 mm (Alastair J. M. Key and Lycett
2014) become markedly inefficient for basic cutting tasks. Similarly,
data from Oldowan replication experiments (Stout et al. 2019) show that
the utility index (flake cutting edge/flake mass1/3) * (1 - exp{[}-0.31
* (flake maximum dimension -- 1.81){]}) developed by Morgan et
al.~(2015) falls off rapidly below 40mm maximum dimension ( Mean Utility
\textless{} 40mm = 0.508; \textgreater=40mm = 0.946; t= 11.99, df = 707,
p \textless{} 0.000). By including weight in our cut-off criteria we
also avoid skewing the flake shape distribution by selectively retaining
long, thin pieces (i.e., MD \textgreater{} 40, weight \textless{} 5g)
while discarding rounder pieces of similar (or greater) weight and area.

For measurement, DP length was defined as the longest axis and width as
the maximum dimension orthogonal to length. Thickness was defined as the
maximum dimension orthogonal to the plane formed by L and W and was
measured using calipers. L, W, and plan-view area measurements were
taken from photographs captured using a Canon Rebel T3i fitted with a 60
mm macro lens and attached to a photographic stand with adjustable upper
and lower light fittings. The camera was positioned directly above the
flakes and kept at a constant height. DPs were positioned irrespective
of any technological features so that the longest axis was vertical, and
the wider end was placed toward the bottom of the photograph.

Photographs were post-processed using Equalight software to adjust for
lens and lighting falloff that result from bending light through a lens
and its aperture which can affect measurements taken from photographs.
Each image was shot with a scale that was then used to rectify the
photograph's pixel scale to a real-world measurement scale in Adobe
Photoshop. Images were converted to binary black and white format and
silhouettes of the tools were extracted in Adobe Photoshop. We then used
a custom ImageJ (Rueden et al. 2017) script (Pargeter, Khreisheh, and
Stout 2019) to measure DP length and take nine width measurements at
10\% increments of length starting at the base of each DP. We used the
built-in ImageJ tool to measure DP area. A ``Proportion Larger DPs'' was
calculated per core as the combined weight of all DPs \textgreater40mm
in maximum dimension and 5g in weight divided by the weight of all DPs.
Higher values show cores with proportionally more large DPs.

@@

To evaluate the association between psychometric, motor-skill, and
training measures and technological outcomes, we adopted an
information-theoretic approach (Burnham and Anderson 2002).
Information-theoretic approaches provide methods for model selection
using all possible combinations of variables while avoiding problems
associated with significance-threshold stepwise selection. We used the
corrected Akaike information criterion (AICc) to rate each possible
combination of predictors on the balance between goodness of fit
(likelihood of the data given the model) and parsimony (number of
parameters). The AICc consists of the log likelihood (i.e., how well
does the model fit the data?) and a penalty term for the number of
parameters that must be estimated in the model (i.e., how parsimonious
is the model?), with a correction for small sample sizes (AICc converges
to the standard AIC at large samples). A lower AICc indicates a more
generalizable model and we used it to compare and rank various possible
models. Each analysis begins with a full model that includes all
predictors of interest. All possible combinations of predictors are then
fit, and the resulting models are ranked and weighted based on their
AICc. The ``best'' model is chosen because it has the lowest AICc score.
